---
title: 'Mini Project: World Population'
author: "Taught by Saghir Bashir (ilustat.com)"
date: "20 May 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  dev = "png",
  fig.ext = ".png")
```

## How to complete this Mini Project

  1. Please read through **ALL** the text below and follow the instructions.
  2. Following the instructions complete and run **R code chunks**.
  3. Check that the **R code chunks** are error free  by clicking the play button.
  4. When you have **completed** the mini project and it is error free you should be able
  to **knit** it into **HTML**.
  5. Take your time and work with your partner.




## World Bank Data: World Population

The World Population data was taken from the World Bank data using the `wbstats` package.
```{r wbstats_Data, echo=TRUE, eval=FALSE}
# library(wbstats)
# To see the orginal male and female data use:
# popF <- wb(indicator = "SP.POP.TOTL.FE.IN")
# popM <- wb(indicator = "SP.POP.TOTL.MA.IN")
```

Note the data were combined and a new CSV file created for this mini project.

*Source:* [World Bank](http://data.worldbank.org/)

## Questions

From the World Population data we would like to answer the following questions:

  1. What are the lowest 10 proportion for females and males by Country and year?
  2. What are the highest 10 proportion for females and males by Country and year?
  3. What are the Portuguese trends for size and percent? 
  4. How does Portugal compare to other European Countries?



## Load Libraries first

Don't forget to load the libraries first:
```{r loadLibs}
# Loading the tidyverse packages
library(tidyverse)
# We also need package stringr later
library(stringr)
```

You must do this each time you start R.

## World Population Original CSV Data

A modified version of the World Population data is present in the file "World_Popn_MF.csv". Read it in using
`read_csv()` from the `readr` package.

```{r WPop_CSV_Data, echo=TRUE, eval=TRUE}
# The World pop
WpopCSV <- "World_Popn_MF.csv"
Wpopn <- read_csv(WpopCSV)
```

### World Population Tidy Data Example

We want to end up with a tidy dataset with all the countries like
```{r Wpop_Tidy_Data}
PopnMF_eg <- read_rds("popnMF_eg.rds")
glimpse(PopnMF_eg)
```

***Exercise***

With your partner look at the datasets (tibbles) `Wpopn` and `PopnMF_eg`.

  - Discuss the structure and content of both datasets. 
  - What do the variables (columns) in `Wpopn` represent?
  - Can the questions be answered with this data?
  - How can you tidy the dataset `Wpopn` in the structure of `PopnMF_eg`?

***Hints***
  - You could use `glimpse()` or `names()` functions on the `Wpopn` and `PopnMF_eg` datasets.
  - You could open the `Wpopn` but it may be presented a bit truncated as there are too many variables to display.




## World Population Creating Tidy Data

To create a tidy version (called `PopnMF`) of the `Wpopn` dataset use the following steps to fill 
in the missing code below:

  1. Put the populations totals into a variable called `Popn` and call the `key` variable "SexYear".
  2. Create variable `Sex` as the first character of variable `SexYear`. [COMPLETED]
  3. Create variable `Year` as numeric part of variable `SexYear`.  [COMPLETED]
  4. Remove variable `SexYear`
  5. Spread the data by `Sex`
  6. Remove any missing "NA" values for both `PopnF` and `PopnM`. [COMPLETED]
  7. Sort (arrange) the data by `Country` and `Year`



```{r Wpopn_tidy}
#--# Fill in all the functions below
PopnMF <- Wpopn %>%
          gather(key =    , value =    , -Country) %>%
          mutate(Sex = str_sub(SexYear, start = 1, end = 5)) %>%
          mutate(Year = parse_number(SexYear)) %>%
          select(    ) %>%
          spread(key =     , value =    ) %>%
          filter(!is.na(PopnF) & !is.na(PopnM)) %>%
          arrange(  )

```



## Answering the Questions

###  1. What are the lowest and highest 10 proportion for females by country and year?


We will create a new dataset called `Popn_all` from dataset `PopnMF` with two new variables 
for the proportions of females and males.
```{r lpropF}
# Solution given 
Popn_all <- PopnMF %>%
                 mutate(PropF = PopnF / (PopnF + PopnM)) %>%
  #--# We add it now as it will be used for Question 2 below.
  #--# Fill in the `mutate` to calclute the proportion for men.
                 mutate(PropM =  ) %>%
                 arrange(PropF, Country, Year)
                 
# Printing the first 10 observations.
Popn_all %>%  head(10) %>%  print
```

***Exercise*** 

What are the 10 highest?
```{r lhpropM}
#--# Method 1: Use the `tail()` function instead of `head()` function.
Popn_all %>%

  
  
#--# Method 2: Use the `arrange()` with a descending (`desc()`) sort.
Popn_all %>% arrange (  ) %>%


```




###  2. What are the lowest and highest 10 proportion for males by country and year?

Using the dataset `Popn_all` do the same for male proportions (`PropM`).

***Hint*** Don't forget to sort (`arrange()`) the data first for males.

```{r hpropF}
#--# Complete to show the lowest 10 proportions for males.
Popn_all %>% arrange (  ) %>%


# Show the highest 10 proportions for males.
Popn_all %>% tail(10) %>% print


```

**Discuss with your partner**

  - What is the relationship between Questions 1 and 2?




###  3. What are the Portuguese trends for size and percent? 

Create dataset called `PopnPT` with only the Portuguese data. Look at this
new dataset.
```{r Portugal_Trends}
#--# C
PopnPT <- Popn_all %>%
  #--# To do: Select only the Portuguese data.
          filter(  )
```


***Graphs***

A couple of plots for females to look at the trends
```{r plot_Females}
# Proportion of females over time.
PopnPT %>% ggplot(aes(Year , PropF)) + 
           geom_line() +
           labs(title = "Portuguese Population") + 
           ylab("Proportion - Females") +
           scale_y_continuous(limits = c(0, 1))

# Number of females over time.
PopnPT %>% ggplot(aes(Year , PopnF)) + 
           geom_line() +
           labs(title = "Portuguese Population") + 
           ylab("Number - Females") 
```

***Exercise***

Make the two plots above for males.
```{r plot_males}
# Proportion of males over time.


# Number of males over time.


```

**Discuss** the trends with your partner.


###  4. How does Portugal compare to other European Countries?

Create dataset called `PopnEU` with the Portuguese data and **four** other EU countries 
of your choice. Your objective is to compare the population trends in all five countries.

This is a free exercise for you to use a mix of what you have learnt and *your* ideas.

***ENJOY!***

```{r PTvsEU}
#--# You are free to decide how to compare :o)


```




## Knit this document.

Once completed and working, click on the "Knit" button above. 

















