---
title: 'Exercises: Women in Parliament'
author: "Taught by Saghir Bashir (ilustat.com)"
date: "20 May 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  dev = "png",
  fig.ext = ".png")
```

## Women in Parliament World Bank Data

**Proportion of seats held by women in national parliaments (%)**

*Source:* [World Bank](http://data.worldbank.org/indicator/SG.GEN.PARL.ZS)

## Questions

From the Women in Parliament data we would like to answer the following questions:

  - How does Portugal compare in Europe?
  - For each female in parliament how many men are there?
  - Which countries have no women in Parliament by year?






## WiP Original Data

The data downloaded from the above World Bank link is presented in `WiP_Orig`. Run the following
chunk and then open it to look at.

```{r WiP_Orig_Data, echo=TRUE, eval=TRUE, message=FALSE}
# We must load the tidyverse package first
library(tidyverse)

# Opening the original data in R 
# Source CSV file: API_SG.GEN.PARL.ZS_DS2_en_csv_v2.csv
WiP_Orig  <- readRDS("WiP_Orig.rds")
```

#### Exercise - Original Data

Open `WiP_Orig` from the *`Environment`* tab (top right in RStudio) or type `View(WiP_Orig)` in the *`Console`* (bottom left in RStudio).

  - Discuss the structure and content with your partner
  - Can the questions be answered with this data?
  - How would you make the data “tidy”?

***Hints***

  - What problems can you foresee with the variable names?
  - What is Variable X62? (Hint 2: Look at source CSV file `header` line after `2016`.)
  - Are the variables correctly identified?
  - Can variables be removed?

```{r lookOrigData}
glimpse(WiP_Orig)
```






## WiP Tidy Data

For the next exercises we will work with a tidy version of `WiP_Orig` dataset. Later we will come back to create it from the original CSV file.

```{r WiP_Tidy_Data, echo=TRUE, eval=TRUE, message=FALSE}
# Tidy version of WiP_Orig
WiP_Tidy  <- readRDS("WiP_Tidy.rds")
```

Have a look at the `WiP_Tidy` dataset. Perhaps try:
```{r WiP_Tidy_look, echo=TRUE, eval=TRUE, message=FALSE}
glimpse(WiP_Tidy)
# or
names(WiP_Tidy)
```

### Keep Selected Variable - dplyr::select()

Using the `select()` function from `dplyr` there are two ways to keep the variable of key interest namely: "Country", "Year" and "Fip.pct". The first method is given what is the second?
```{r dplyr_select}
# Method 1
WiP_s1 <- select(WiP_Tidy, Country_Name, Year, Fip.pct)
glimpse(WiP_s1)

#--# Method 2 - Fill in the select() function to give the same in a different way.
#--# Hint: Use "-variable"
WiP_s2 <- select( )
glimpse(WiP_s2)
```

### Create New Variable - dplyr::mutate()

Using the `mutate()` function from `dplyr` create the following two variables:

  - % Men in Parliament (Mip.pct) as: 100 - % Females in Parliament
  - Male to female ratio as: (% Men in Parliament) divided by (% Females in Parliament)

```{r dplyr_select}
# % Men in Parliament (Mip.pct)
WiP_m1 <- mutate(WiP_Tidy, Mip.pct = 100 - Fip.pct)
glimpse(WiP_m1)

#--# Method 2 - Fill in the mutate( ) to calculate the Male to Female ratio called `MF_ratio`
WiP_m2 <- mutate( )
glimpse(WiP_m2)
```

### Create New Variable - dplyr::rename()

Using the `rename()` function from `dplyr` rename  `Country_Name` to `Country`:
```{r dplyrs_rename}
#--# Fill in the rename( ) function to rename  `Country_Name` to `Country`
WiP_r1 <- rename( )  
names(WiP_r1)
```

### Create New Variable - dplyr::filter()

Using the `filter()` function from `dplyr`: 
```{r WiP_pt}
# only keep the observations (rows) for Portugual data from 2010 onwards
WiP_pt1 <- filter(WiP_Tidy, Country_Name == "Portugal" & Year >= 2010)
WiP_pt1

# Method 1: Keep the observations (rows) for Portugual for 2000 and 2016 
WiP_pt2 <- filter(WiP_Tidy, Country_Name == "Portugal" & (Year == 2010 | Year ==2016))
WiP_pt2
# Method 2: Keep the observations (rows) for Portugual for 2000 and 2016 
WiP_pt3 <- filter(WiP_Tidy, Country_Name == "Portugal" & Year %in% c(2000, 2016))
WiP_pt3

#--# Fill in the filter function to keep all data for "Angola", "Brazil" and "Portugal"
WiP_pt4 <- filter( )
WiP_pt4

#--# Fill in the filter function to keep all data for "Angola", "Brazil" and "Portugal"
#--# where % females in parliament is less than 10%.
WiP_pt5 <- filter( )
WiP_pt5
```

### Create New Variable - dplyr::arrange()

Using the `arrange()` function from `dplyr`: 

```{r dplyr_arrange}
# Arrange the data by country and descending year
WiP_a1 <- arrange(WiP_Tidy, Country_Name, desc(Year))
WiP_a1

#--# Fill in the arrange function below to sort the data by descending % females in parliament
#--# (highest percent of females first), Country_Name and Year
WiP_a2 <- arrange( )
WiP_a2

#--# Fill in the arrange function below to sort the data by descending % females in parliament
#--# (highest percent of females first), Year and Country_Name -- VARIABLE ORDER CHANGED from WiP_a2
WiP_a3 <- arrange( )
WiP_a3
```

### Pipes - %>% from `magrittr`

We will use the pipe operator `%>%` to do multiple steps.

Create `WiP_p1` as:

  1. Keep the observations for Countries "Cuba", "Peru" and "Brazil"
  2. then keep the variables `Country_Name`, `Year` and `Fip:pct`
  3. then rename `Country_Name` to `Country`
  4. then create variable % Males in Parliament (Mip.pct)
  5. then arrange the data by Country, descending Year

```{r pipe_example}
# Hint: read `%>%` as "then" do.
WiP_p1 <- WiP_Tidy %>%
          filter(Country_Name %in% c("Cuba", "Peru", "Brazil")) %>%
          select(Country_Name, Year, Fip.pct) %>%
          rename(Country = Country_Name) %>%
          mutate(Mip.pct = 100 - Fip.pct) %>%
          arrange(Country, desc(Year))
WiP_p1
```




#### Exercise - Tidy Data

Create `WiP_TD` as:

  1. Keep the variables `Country_Name`, `Year` and `Fip:pct`
  2. then rename `Country_Name` to `Country`
  3. then create variable % Males in Parliament (Mip.pct)
  4. then the male to female ratio (MF_ratio)
  5. then arrange the data by Country, Year

```{r WiP_Tidy_Exercise}
#--# Complete the code below to create `WiP_TD` as described above using `%>%` (pipe operator)
WiP_TD %>% WiP_Tidy %>%
  
  
glimpse(WiP_TD)
```

Finally open the dataset and look through it to see it makes sense.



### Spreading the data - tidyr::spread()

We will use the `spread()` function from the `tidyr` to show data from Angola, Portugal and Brazil
side by side.
```{r tidyr_spread}
WiP_ABP <- WiP_Tidy %>%
           select(Country_Name, Year, Fip.pct) %>%
           filter(Country_Name %in% c("Angola", "Brazil", "Portugal")) %>%
           spread(key = Country_Name, value = Fip.pct)
glimpse(WiP_ABP)
```

Using the `gather()` function from the `tidyr` put the data back into the original format.
```{r tidyr_gather}
#--# Fill in the filter function to put the Country data back into one column.
# Note we call the old "Country_Name" - "Country" now.
WiP_gat <- WiP_ABP %>%
           gather(Angola:Portugal, key = Country, value = Fip.pct)

glimpse(WiP_gat)
```





## Knit this document.

Once completed and working, click on the "Knit" button above. 

